Behavioural Analysis of Independent Value-Based Learning in Non-cooperative Games

### 2025-06-07 paper/paper.tex
paper v8

\begin{abstract}
Multi-agent reinforcement learning has received a lot of updates in cooperative games recently. However, research in general non-cooperative games is currently lagging behind. Independent value-based algorithms such as independent Q-learning have been one of the most widely studied methods in the field due to their simplicity and scalability. Some research has shown that independent algorithms can heavily outperform other approaches in some cooperative scenarios. In this paper, we discuss the feasibility of these algorithms in non-cooperative settings. As a result, we explain the conditions that a game must satisfy for the algorithms to work. We further propose the game Food Chain that simulates an ecosystem and is designed to challenge the algorithms in a way aligned with our analysis. Our results show that independent value-based learning algorithms are guaranteed to fail in some non-cooperative games but are more likely to work in cooperative games.
\end{abstract}

@@

\subsection{MARL Algorithms}
Independent learning (IL) algorithms that are carried over from the traditional single-agent RL domain treat other agents as part of the environment. The agents are essentially unaware of the existence of one another. IL algorithms are popular for handling many complex problems \cite{papoudakis2020benchmarking, gupta2017cooperative, de2020independent, palmer2020independent}. Among these, independent value-based learning (IVBL) algorithms, such as the classic independent Q-learning (IQL) \cite{tan1993multi}, learn the expected long-term cumulative return (the Q-value) of each action at each state through sampling. IVBL algorithms including IQL are often considered as baseline and natural choice in a wide range of MARL studies because of their simplicity and scalability. For example, they were studied in various cooperative environments \cite{foerster2017stabilising, omidshafiei2017deep, palmer2017lenient, palmer2018negative}. The problem is that IVBL agents only help produce the average reward of each action for each agent, but do not pay attention on what the action-selection policy. This is not a problem in single-agent or cooperative environments as we can simply pick the action with the highest return. Some studies addressed this issue in non-cooperative games by simply choosing the action with the highest empirical return \cite{bjornsson2009cadiaplayer, jiang2018q, kopacz2023evaluating} and observed reasonably good outcome. Kopacz et al \cite{kopacz2023evaluating} has examined IVBL in the non-cooperative game predator-and-prey. Both the algorithm and the game are closely related to our work. Other researchers \cite{qu2020distributed} tried to address this problem using the softmax function given in Equation~(\ref{eq:softmax}) to calculate a policy based on the payoff. Softmax takes a vector as input and computes a probability distribution over the original vector entries, such that the higher the original value, the higher the probability.
\begin{equation}
    \textbf{softmax}([x_1, ..., x_n]) = [\frac{e^{x_1}}{\sum_{i}{e^{x_i}}}, ..., \frac{e^{x_n}}{\sum_{i}{e^{x_i}}}]
    \label{eq:softmax}
\end{equation}
The latter approach emphasis the action with higher empirical return. However, these approaches have a huge problem that guarantees failure of learning in a wide range of games. We will take a closer look in the analysis section.

@@

\begin{figure}
\centering
\subfloat[A Food Web]{
    \resizebox*{0.5\textwidth}{!}{
        \includegraphics{images/food_web.png}
    }
}
\subfloat[The Food Chain Game]{
    \resizebox*{0.5\textwidth}{!}{
        \includegraphics{images/food_chain_game.png}
    }
}
\caption{(a), A food web represents a simple ecosystem. The food web is a cluster of multiple food chains showing what-eats-what. The arrows point in the direction of the energy flow. (b), In the food chain game, an agent represents an individual animal of some type. They navigate on a playground trying to feed themselves while avoiding enemies.}
\label{fig:food_chain_game}
\end{figure}

@@

\section{Experimental Evaluation}
Too see how Food Chain can turn on/off the necessary condition (\ref{eq:necessary_condition}) for IVBL to converge, we design two scenarios. The key difference between the two is that one satisfies the condition (\ref{eq:necessary_condition}) and the other does not. We define a simple food web consisting of frogs and hoppers. Hoppers (represented by h) can consume vegetation (drawn as *) and will be eaten by frogs (f). Both scenarios will turn off random food generation.

\begin{figure}
\centering
    \begin{minipage}{0.48\textwidth}
        \centering
        \includegraphics[width=0.931\textwidth]{images/biased_rps.png}
        \caption{The initial state of Biased RPS in Food Chain. The frog (f) is trying to capture the hopper (h). The only NE in the game contains non-uniformly random mixed policies.}
        \label{fig:biased_rps_in_food_chain}
    \end{minipage}\hfill
    \begin{minipage}{0.48\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{images/tree_maze_chase.png}
        \caption{The initial state of Tree Maze Chase in Food Chain. The frog (f) is chasing the hooper (h) in the maze. There is an optimal pure (deterministic) policy for both players.}
        \label{fig:tree_maze_chase}
    \end{minipage}
\end{figure}

\subsection{Biased Rock-Paper-Scissors in Food Chain}
This case violates condition (\ref{eq:necessary_condition}) and forces the agents to address the mixed policy problem. The initial state of the scenario consists of a frog trying to hunt down a hopper as shown in Figure~\ref{fig:biased_rps_in_food_chain}. We can see that in some cases the hopper can escape. If they both pass, they will face the exact same situation again. Figure~\ref{fig:biased_rps_in_food_chain_possible_next_states} in the appendix shows all possible next states. For simplicity, we set the max timestep to 1, so the game ends after one move. The scenario becomes a normal form mini game. To break the symmetry so that condition (\ref{eq:necessary_condition}) is violated, we add food in one cell, making the scenario essentially biased RPS. We also raise the feeding reward for both agents to 0.5 to highlight the asymmetry. The reward matrix is shown in Table~\ref{table:biased_rps_in_food_chain_reward_matrix} in the appendix. We can verify that there is a unique NE in the game.
\begin{equation}
    \pi_h=(P(r)=\frac{1}{3},P(d)=\frac{1}{3},P(p)=\frac{1}{3})
    \label{eq:biased_rps_in_food_chain_hopper_policy}
\end{equation}
\begin{equation}
    \pi_f=(P(l)=\frac{1}{13},P(u)=\frac{6}{13},P(p)=\frac{6}{13})
    \label{eq:biased_rps_in_food_chain_frog_policy}
\end{equation}
The expected reward for the hooper and the frog should be $-7/13 \approx -0.538$ and $1/3 \approx 0.333$, respectively.

In the Food Chain game, because all agents have the same mobility, a single predator can only capture a prey by trapping them in a corner. Thus, any predator, unless working in a team, has to face this subcase at some point during a general Food Chain game played on an open ground. This means that the scenario represents a subgame which will show up frequently in a full game.

\subsection{Tree Maze Chase in Food Chain}
This scenario essentially turns off the mixed policy problem, so condition (\ref{eq:necessary_condition}) is met. We set up the initial state of the game as shown in Figure~\ref{fig:tree_maze_chase} where a frog is chasing a hopper in a maze. This map resembles a tree structure (no cycles) which ensures that there is a pure NE in the game, because there is always an optimal action for both players. The hooper should go for the tunnel with more food and the frog should simply chase the hopper. This means the necessary condition (\ref{eq:necessary_condition}) is satisfied. To make sure that at most two food can be picked up by the hopper, we set the max timestep to 12. The optimal rewards for the hopper and the frog are -0.8 and 1, respectively.

\subsection{Results and Discussion}
In both scenarios, we test the IVBL algorithm IQL and the independent policy-based algorithm IPPO. All experiments were run for 800 iterations. For each iteration, the algorithm collects 640 timesteps as training samples and train the neural network. After each training iteration, 50 evaluation games are run during the evaluation phase. The plots show the average reward for each agent during the evaluation phase. Each experiment is run 20 times to mitigate the effect of randomness and the graphs show the boots trapped 95\% confidence interval. In total, each data point on the graph is the average result of $20 \times 50 = 1000$ evaluation games. More details of the parameters for the experiments can be found in Table~\ref{table:experiments_parameters} in the appendix.

@@

\begin{figure}
\centering
\subfloat[Biased RPS Hopper Rewards]{
    \resizebox*{0.46\textwidth}{!}{
        \includesvg{plots/food_chain_biased_rps_hopper_return.svg}
    }
}
\subfloat[Biased RPS Frog Rewards]{
    \resizebox*{0.46\textwidth}{!}{
        \includesvg{plots/food_chain_biased_rps_frog_return.svg}
    }
}
\caption{Biased RPS. IVBL (IQL) shows significantly more instability and does not converge to the NE.}
\label{fig:biased_rps_in_food_chain_rewards}
\end{figure}

\begin{figure}
\centering
\subfloat[Tree Maze Chase Hopper Rewards]{
    \resizebox*{0.46\textwidth}{!}{
        \includesvg{plots/food_chain_tree_maze_chase_hopper_return.svg}
    }
}
\subfloat[Tree Maze Chase Frog Rewards]{
    \resizebox*{0.46\textwidth}{!}{
        \includesvg{plots/food_chain_tree_maze_chase_frog_return.svg}
    }
}
\caption{Tree Maze Chase. Both algorithms quickly learned to play the game and tend to stay in the optimal reward range. Note that increasing frog reward reduces the hopper reward as the game is non-cooperative.}
\label{fig:tree_maze_chase_rewards}
\end{figure}

\section{Conclusion}
In this work, we have analysed the stability of IVBL algorithms in non-cooperative games and tested their performance in specific game scenarios. A necessary condition (\ref{eq:necessary_condition}) for IVBL algorithms to converge to a NE in finite games was identified. The condition requires the game to have a NE where every agent's policy evenly chooses their worthy actions (actions that are played with a positive probability). This explains why we may observe satisfying results when applying IVBL to games with pure NEs including cooperative games. We further proposed the game Food Chain which has certain traits that can verify our analysis. The test results show that IVBL algorithms perform poorly in scenarios where the condition (\ref{eq:necessary_condition}) is violated. Our work is limited to small two-player scenarios in the Food Chain game. There are still interesting characteristics of the game that may be explored in the future. Such as the extension in both the number of agents and the number of species in the food web.

@@

\section{Biased RPS in Food Chain}
Figure~\ref{fig:biased_rps_in_food_chain_possible_next_states} shows all the possible next state of the Biased RPS scenario in Food Chain. Table~\ref{table:biased_rps_in_food_chain_reward_matrix} describes the reward matrix of the scenario (note that if both agents try to move into the food tile, the hopper only has a 50\% chance to get food).

@@

\section{Experiments Parameters}
\begin{table}
\tbl{Experiments Parameters}{
    \begin{tabular}{lcc}
        \toprule
        \multicolumn{3}{l}{game parameters} \\
        \midrule
        & Biased RPS & Tree Maze Chase \\
        \cmidrule{2-3}
        size of the map & $2 \times 2$ & $5 \times 5$ \\
        max timestep & 1 & 12 \\
        feeding reward & 0.5 & 0.1 \\
        death reward & \multicolumn{2}{c}{-1} \\
        food spawn chance & \multicolumn{2}{c}{0} \\
        number of agents & \multicolumn{2}{c}{2} \\
        agent types & \multicolumn{2}{c}{2} \\
        \midrule
        \multicolumn{3}{l}{experiment parameters} \\
        \midrule
        repetitions of each experiment & \multicolumn{2}{c}{20} \\
        number of iterations & \multicolumn{2}{c}{800} \\
        evaluation games after each iteration & \multicolumn{2}{c}{50} \\
        \midrule
        \multicolumn{3}{l}{training hyper parameters} \\
        \midrule
        & IPPO & IQL \\
        \cmidrule{2-3}
        on-policy minibatch optimization iterations\textsuperscript{a} & 10 & \\
        off-policy optimizer steps\textsuperscript{b} & & 100 \\
        off-policy memory buffer size & & 20,000 \\
        frames collected for each iteration & \multicolumn{2}{c}{640} \\
        minibatch size & \multicolumn{2}{c}{32} \\
        discount factor $\gamma$ & \multicolumn{2}{c}{0.99} \\
        learning rate $\alpha$ & \multicolumn{2}{c}{$5 \times 10^{-5}$} \\
        the $\varepsilon$ parameter of the adam optimizer & \multicolumn{2}{c}{$1 \times 10^{-6}$} \\
        \bottomrule
    \end{tabular}
}
\tabnote{\textsuperscript{a}The number of times collected frames will be split into minibatches and trained. \\ \textsuperscript{b}The number of times a minibatch will be sampled from the memory buffer and trained over.}
\label{table:experiments_parameters}
\end{table}

>>>>

\begin{abstract}
Multi-agent reinforcement learning has received increased attention in cooperative games recently. However, research in general non-cooperative games is currently lagging behind. Independent value-based algorithms, such as independent Q-learning, have demonstrated simple implementation and versatility in various contexts. Previous research has shown that independent algorithms can heavily outperform other approaches in some cooperative scenarios. In this paper, we study the behaviour of these algorithms in non-cooperative settings. As a result, we explain the conditions that a game must satisfy for the algorithms to work. We further propose the game Food Chain that simulates an ecosystem and is designed to challenge the algorithms in a way aligned with our analysis. We collect empirical data on the performance of selected algorithms from experiments based on the proposed game. Our results show that independent value-based learning algorithms
can converge to Nash equilibrium in non-cooperative environments, only when the Nash equilibrium consists of uniformly random policies over the feasible actions.
%are guaranteed to fail in some non-cooperative games but are more likely to work in cooperative games.
\end{abstract}

@@

\subsection{MARL Algorithms}
Independent learning (IL) algorithms that are carried over from the traditional single-agent RL domain treat other agents as part of the environment. The agents are essentially unaware of the existence of one another. IL algorithms are popular for handling many complex problems \cite{papoudakis2020benchmarking, gupta2017cooperative, de2020independent, palmer2020independent}. Among these, independent value-based learning (IVBL) algorithms, such as the classic independent Q-learning (IQL) \cite{tan1993multi}, learn the expected long-term cumulative return (the Q-value) of each action at each state through sampling. IVBL algorithms including IQL are often considered as baseline and natural choice in a wide range of MARL studies because of their simplicity and scalability. For example, they were studied in various cooperative environments \cite{foerster2017stabilising, omidshafiei2017deep, palmer2017lenient, palmer2018negative}. The problem is that IVBL agents only help produce the average reward of each action for each agent, but do not pay attention on what the action-selection policy. This is not a problem in single-agent or cooperative environments as we can simply pick the action with the highest return. Some studies addressed this issue in non-cooperative games by simply choosing the action with the highest empirical return \cite{bjornsson2009cadiaplayer, jiang2018q, kopacz2023evaluating} and observed reasonably good outcome. Kopacz et al \cite{kopacz2023evaluating} has examined IVBL in the non-cooperative game predator-and-prey. Both the algorithm and the game are closely related to our work. Other researchers \cite{qu2020distributed} tried to address this problem using the softmax function given in Equation~(\ref{eq:softmax}) to calculate a policy based on the payoff. Softmax takes a vector as input and computes a probability distribution over the original vector entries, such that the higher the original value, the higher the probability.
\begin{equation}
    \textbf{softmax}([x_1, ..., x_n]) = [\frac{e^{x_1}}{\sum_{i}{e^{x_i}}}, ..., \frac{e^{x_n}}{\sum_{i}{e^{x_i}}}]
    \label{eq:softmax}
\end{equation}
The latter approach emphasis the action with higher empirical return. However, these approaches have a problem that guarantees failure of learning in a wide range of games. We will take a closer look in the analysis section.

@@

\begin{figure}
\centering
\subfloat[A Food Web]{
    \resizebox*{0.48\textwidth}{!}{
        \includegraphics{images/food_web.png}
    }
}
\subfloat[The Food Chain Game]{
    \resizebox*{0.48\textwidth}{!}{
        \includegraphics{images/food_chain_game.png}
    }
}
\caption{(a), A food web represents a simple ecosystem. The food web is a cluster of multiple food chains showing what-eats-what. The arrows point in the direction of the energy flow. (b), In the Food Chain game, an agent represents an individual animal of some type. They navigate on a playground trying to feed themselves while avoiding enemies.}
\label{fig:food_chain_game}
\end{figure}

@@

\section{Experimental Evaluation}
Too see how Food Chain can turn on/off the necessary condition (\ref{eq:necessary_condition}) for IVBL to converge, we design two scenarios. The key difference between the two is that one satisfies the condition (\ref{eq:necessary_condition}) and the other does not. We define a simple food web consisting of frogs and hoppers. Hoppers (represented by h) can consume vegetation (drawn as *) and will be eaten by frogs (f). Both scenarios turns off vegetation generation.

\begin{figure}
\centering
    \begin{minipage}{0.48\textwidth}
        \centering
        \includegraphics[width=0.931\textwidth]{images/biased_rps.png}
        \caption{The initial state of Biased RPS in Food Chain. The frog (f) is trying to capture the hopper (h). The only NE in the game contains non-uniformly random mixed policies.}
        \label{fig:biased_rps_in_food_chain}
    \end{minipage}\hfill
    \begin{minipage}{0.48\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{images/tree_maze_chase.png}
        \caption{The initial state of Tree Maze Chase in Food Chain. The frog (f) is chasing the hopper (h) in the maze. There is an optimal pure (deterministic) policy for both players.}
        \label{fig:tree_maze_chase}
    \end{minipage}
\end{figure}

\subsection{Biased Rock-Paper-Scissors in Food Chain}
This case violates condition (\ref{eq:necessary_condition}). The initial state of the scenario consists of a frog trying to hunt down a hopper as shown in Figure~\ref{fig:biased_rps_in_food_chain}. We can see that in some cases the hopper can escape. If they both pass, they will face the exact same situation again. Figure~\ref{fig:biased_rps_in_food_chain_possible_next_states} in the appendix shows all possible next states. We set the max timestep to 1, so the scenario becomes a normal form game. To break the symmetry so that condition (\ref{eq:necessary_condition}) is violated, we add vegetation in one cell, making the scenario essentially biased RPS. We also raise the feeding reward for both agents to 0.5 to highlight the asymmetry. The reward matrix is shown in Table~\ref{table:biased_rps_in_food_chain_reward_matrix} in the appendix. We can verify that there is a unique NE in the game.
\begin{equation}
    \pi_h=(P(r)=\frac{1}{3},P(d)=\frac{1}{3},P(p)=\frac{1}{3})
    \label{eq:biased_rps_in_food_chain_hopper_policy}
\end{equation}
\begin{equation}
    \pi_f=(P(l)=\frac{1}{13},P(u)=\frac{6}{13},P(p)=\frac{6}{13})
    \label{eq:biased_rps_in_food_chain_frog_policy}
\end{equation}
The expected reward for the hopper and the frog should be $-7/13 \approx -0.538$ and $1/3 \approx 0.333$, respectively.

In the Food Chain game, because all agents have the same mobility, a predator can only capture a prey by trapping them in a corner. Thus, any predator, unless working in a team, has to face this subcase at some point during a general Food Chain game played on an open ground. The scenario represents a subgame which shows up frequently in a full game.

\subsection{Tree Maze Chase in Food Chain}
This scenario essentially turns off the mixed policy problem. We set up the initial state as shown in Figure~\ref{fig:tree_maze_chase} where a frog is chasing a hopper in a maze. This map resembles a tree structure (no cycles) which creates a pure NE, because there is always an optimal action. The hopper should go for the tunnel with more vegetation and the frog should simply chase the hopper. The necessary condition (\ref{eq:necessary_condition}) is satisfied. To make sure that at most two vegetations can be picked up by the hopper, we set the max timestep to 12. The optimal rewards for the hopper and the frog are -0.8 and 1, respectively.

\subsection{Results and Discussion}
We tested the IVBL algorithm IQL and the independent policy-based algorithm IPPO in both scenarios. All experiments ran for 800 iterations. For each iteration, the algorithm collects 640 timesteps as training samples and train the neural network before 50 evaluation games were run. The plots show the average reward for each agent during the evaluation phase. Each experiment was run 20 times to mitigate the effect of randomness and the graphs show the boots trapped 95\% confidence interval. More details of the parameters for the experiments can be found in Table~\ref{table:experiments_parameters} in the appendix.

\begin{figure}
\centering
\subfloat[Biased RPS Hopper Rewards]{
    \resizebox*{0.48\textwidth}{!}{
        \includesvg{plots/food_chain_biased_rps_hopper_return.svg}
    }
}
\subfloat[Biased RPS Frog Rewards]{
    \resizebox*{0.48\textwidth}{!}{
        \includesvg{plots/food_chain_biased_rps_frog_return.svg}
    }
}
\caption{Biased RPS. IVBL (IQL) shows significantly more instability and does not converge to the NE.}
\label{fig:biased_rps_in_food_chain_rewards}
\end{figure}

\begin{figure}
\centering
\subfloat[Tree Maze Chase Hopper Rewards]{
    \resizebox*{0.48\textwidth}{!}{
        \includesvg{plots/food_chain_tree_maze_chase_hopper_return.svg}
    }
}
\subfloat[Tree Maze Chase Frog Rewards]{
    \resizebox*{0.48\textwidth}{!}{
        \includesvg{plots/food_chain_tree_maze_chase_frog_return.svg}
    }
}
\caption{Tree Maze Chase. Both algorithms quickly learned to play the game and tend to stay in the optimal reward range. Note that increasing frog reward reduces the hopper reward as the game is non-cooperative.}
\label{fig:tree_maze_chase_rewards}
\end{figure}

@@

\section{Conclusion}
In this work, we have analysed the stability of independent value-based learning (IBVL) algorithms in non-cooperative games and tested their performance in specific game scenarios. By means of proof by contradiction and the design of experiments based on the proposed Food Chain game, we found that the studied algorithms only converge to Nash equilibrium (NE) in non-cooperative settings when every agent's policy evenly chooses the next action among their worthy actions (actions that are played with a positive probability). This explains why promising results are achieved when applying IVBL to games with pure NE, including cooperative games.

This work also proposes the Food Chain game which has certain traits ---such as ease of navigation, focus on interactions, and extendability, among others--- that are instrumental for the analysis of the stability in non-cooperative settings. The test results show that IVBL algorithms perform poorly in scenarios where the identified condition is violated. Our work is limited to small two-player scenarios in the Food Chain game. There are still interesting characteristics of the game that will be explored in the future. Such as the extension in both the number of agents and the number of species on the food web.

@@

\section{Biased RPS in Food Chain}
Figure~\ref{fig:biased_rps_in_food_chain_possible_next_states} shows all the possible next state of the Biased RPS scenario in Food Chain. Table~\ref{table:biased_rps_in_food_chain_reward_matrix} describes the reward matrix of the scenario (note that if both agents try to move into the vegetation tile, the hopper only has a 50\% chance to get the vegetation).

@@

\section{Experiments Parameters}
\begin{table}
\tbl{Experiments Parameters}{
    \begin{tabular}{lcc}
        \toprule
        \multicolumn{3}{l}{game parameters} \\
        \midrule
        & Biased RPS & Tree Maze Chase \\
        \cmidrule{2-3}
        size of the map & $2 \times 2$ & $5 \times 5$ \\
        max timestep & 1 & 12 \\
        feeding reward & 0.5 & 0.1 \\
        death reward & \multicolumn{2}{c}{-1} \\
        vegetation spawn chance & \multicolumn{2}{c}{0} \\
        number of agents & \multicolumn{2}{c}{2} \\
        agent types & \multicolumn{2}{c}{2} \\
        \midrule
        \multicolumn{3}{l}{experiment parameters} \\
        \midrule
        repetitions of each experiment & \multicolumn{2}{c}{20} \\
        number of iterations & \multicolumn{2}{c}{800} \\
        evaluation games after each iteration & \multicolumn{2}{c}{50} \\
        \midrule
        \multicolumn{3}{l}{training hyper parameters} \\
        \midrule
        & IPPO & IQL \\
        \cmidrule{2-3}
        on-policy minibatch optimization iterations\textsuperscript{a} & 10 & \\
        off-policy optimizer steps\textsuperscript{b} & & 100 \\
        off-policy memory buffer size & & 20,000 \\
        frames collected for each iteration & \multicolumn{2}{c}{640} \\
        minibatch size & \multicolumn{2}{c}{32} \\
        discount factor $\gamma$ & \multicolumn{2}{c}{0.99} \\
        learning rate $\alpha$ & \multicolumn{2}{c}{$5 \times 10^{-5}$} \\
        the $\varepsilon$ parameter of the adam optimizer & \multicolumn{2}{c}{$1 \times 10^{-6}$} \\
        \bottomrule
    \end{tabular}
}
\tabnote{\textsuperscript{a}The number of times collected frames will be split into minibatches and trained. \\ \textsuperscript{b}The number of times a minibatch will be sampled from the memory buffer and trained over.}
\label{table:experiments_parameters}
\end{table}
